---
title: "Homefield Adantage Stats"
author: "Helen Threlkeld and Molly Long"
date: "1/21/2021"
output: word_document
---
```{r}
data1 <- read.csv("MycoDB_version4.csv", header = T)
data2<- subset(data1, select = c(PlantFamily,EFFECTSIZE1,NONMYCOCONTROL2,FERTP,FERTN,INOC.COMPLEXITY,STERILIZED,LOCATION,LA_Code,ESTVAR1,ESTVAR3), Rua2016=="YES")
data<-data2[data2$LA_Code %in% c("A","B","C","D","E"), ]
data$PlantFamily <- as.factor(data$PlantFamily)
data$EFFECTSIZE1 <- as.numeric(data$EFFECTSIZE1)
data$NONMYCOCONTROL2 <- as.factor(data$NONMYCOCONTROL2)
data$FERTP <- as.factor(data$FERTP)
data$FERTN <- as.factor(data$FERTN)
data$INOC.COMPLEXITY <- as.factor(data$INOC.COMPLEXITY)
data$STERILIZED <- as.factor(data$STERILIZED)
data$LOCATION <- as.factor(data$LOCATION)
data$LA_Code <- as.factor(data$LA_Code)
#scale(data2$EFFECTSIZE1, )
?scale
#todo: scale numeric variables 
```
###New analysis ANOVA
```{r}
library(car)
library(dplyr)
library(rstatix)
library(tidyverse)
library(ggpubr)

#Check Assumptions 

#Check outliers 
data %>%
 group_by(LA_Code, LOCATION,INOC.COMPLEXITY) %>%
 identify_outliers(EFFECTSIZE1)

#Check normality
model  <- lm(EFFECTSIZE1 ~ LA_Code*LOCATION*INOC.COMPLEXITY, data = data)
# Create a QQ plot of residuals
ggqqplot(residuals(model))
# Compute Shapiro-Wilkes test of normality
shapiro.test(residuals(model))

#QQ Plot by LA_Code and Inoculum complexity 
ggqqplot(data, "EFFECTSIZE1", ggtheme = theme_bw()) +
  facet_grid( LOCATION + INOC.COMPLEXITY ~ LA_Code, labeller = "label_both")

#Homogenetiy variance assumption
leveneTest(data$EFFECTSIZE1 ~ data$LA_Code*data$LOCATION*data$INOC.COMPLEXITY)
#p-val < 0.05 so the variance is homogenous 

#anova test 
#res.aov <- data %>% anova_test(EFFECTSIZE1 ~ LA_Code*LOCATION*INOC.COMPLEXITY)
#summary(res.aov)

aov3 = aov(EFFECTSIZE1 ~ LA_Code*LOCATION*INOC.COMPLEXITY, data = data)
summary(aov3)

TukeyHSD(aov3)
plot(TukeyHSD(aov3))

#The variable that had a significant effect on EFFECT SIZE was LA_Code

#There is significant difference between A&E and C&E, single vs multi inoc, E:Single vs A:Whole, C:Lab vs E:Lab, 


#source:https://www.datanovia.com/en/lessons/anova-in-r/#three-way-independent-anova
```
###Qe
##Null Hypothesis: No difference among effect sizes.
#Compares effect sizes. 

###Qm
##Null Hypothesis: Model using predictor variable is no stronger than general model. 
#Similar to X2 test and based on X2 distribution. 

###Planned Contrast test 
#??: Not clear if it is a t-test or something else.]
#Planned contrast is a post-hoc test for an ANOVA. 

###Mean Effect Size 
```{r}
#todo: change y-axis line divisions, planned contrasts 
#Full Data 
library(stats)
library(ggplot2)
by_LA_Code.fds = data %>%
  group_by(LA_Code)

by_LA_Code.summary.fds = by_LA_Code.fds %>% summarise(
  m.fds = mean(EFFECTSIZE1),
  sd.fds = sd(EFFECTSIZE1),
  se.fds = sd(EFFECTSIZE1/sqrt(n())),
  lower.fds = m.fds - se.fds,
  upper.fds = m.fds + se.fds)

#Full data set graph
ggplot(by_LA_Code.summary.fds, aes(LA_Code, m.fds)) + geom_point(size = 3) + geom_errorbar(aes(ymin = lower.fds, ymax= upper.fds), width = 0.1) + xlab("LA_Code") + ylab("Mean Effect Size") + ggtitle("Full Data Set") + ylim(-1,1.5) + geom_hline(yintercept=0, linetype="dashed")

#Single Innocula 
data.single = subset(data, INOC.COMPLEXITY == "Single")
by_LA_Code.single = data.single %>%
  group_by(LA_Code)

by_LA_Code.single.summary = by_LA_Code.single %>% summarise(
  m.single = mean(EFFECTSIZE1),
  sd.single = sd(EFFECTSIZE1), 
  se.single = sd(EFFECTSIZE1/sqrt(n())),
  lower.single = m.single - se.single, 
  upper.single = m.single + se.single)

ggplot(by_LA_Code.single.summary, aes(LA_Code, m.single)) + geom_point(size = 3) + geom_errorbar(aes(ymin = lower.single, ymax= upper.single), width = 0.1) + xlab("LA_Code") + ylab("Mean Effect Size") + ggtitle("Single Innocula") + ylim(-1,1.5) + geom_hline(yintercept=0, linetype="dashed")

#Lab Studies 
data.lab = subset(data, LOCATION == "lab")
by_LA_Code.lab = data.lab %>% 
  group_by(LA_Code)

by_LA_Code.lab.summary = by_LA_Code.lab %>% summarise(
  m.lab = mean(EFFECTSIZE1), 
  sd.lab = sd(EFFECTSIZE1),
  se.lab = sd(EFFECTSIZE1/sqrt(n())),
  lower.lab = m.lab - se.lab, 
  upper.lab = m.lab + se.lab)

ggplot(by_LA_Code.lab.summary, aes(LA_Code, m.lab)) + geom_point(size = 3) + geom_errorbar(aes(ymin = lower.lab, ymax= upper.lab), width = 0.1) + xlab("LA_Code") + ylab("Mean Effect Size") + ggtitle("Lab") + ylim(-1,1.5) + geom_hline(yintercept=0, linetype="dashed")

```

###Weighted Mean Effect Size 
##Fig 2 
```{r}
#Full Data 
library(stats)
library(ggplot2)
by_LA_Code.summary.fds.w = by_LA_Code.fds %>% summarise(
  wm.fds = weighted.mean(EFFECTSIZE1),
  sd.fds = sd(EFFECTSIZE1),
  se.fds = sd(EFFECTSIZE1/sqrt(n())),
  lower.fds.w = wm.fds - se.fds,
  upper.fds.w = wm.fds + se.fds)

#Full data set graph
ggplot(by_LA_Code.summary.fds.w, aes(LA_Code, wm.fds)) + geom_point(size = 3) + geom_errorbar(aes(ymin = lower.fds.w, ymax= upper.fds.w), width = 0.1) + xlab("LA_Code") + ylab("Weighted Mean Effect Size") + ggtitle("Full Data Set") + ylim(-1,1.5) + geom_hline(yintercept=0, linetype="dashed") 
#Single Innocula 

by_LA_Code.single.summary.w = by_LA_Code.single %>% summarise(
  wm.single = mean(EFFECTSIZE1),
  sd.single = sd(EFFECTSIZE1), 
  se.single = sd(EFFECTSIZE1/sqrt(n())),
  lower.single.w = wm.single - se.single, 
  upper.single.w = wm.single + se.single)

ggplot(by_LA_Code.single.summary.w, aes(LA_Code, wm.single)) + geom_point(size = 3) + geom_errorbar(aes(ymin = lower.single.w, ymax= upper.single.w), width = 0.1) + xlab("LA_Code") + ylab("Weighted Mean Effect Size") + ggtitle("Single Innocula") + ylim(-1,1.5) + geom_hline(yintercept=0, linetype="dashed")

#Lab Studies 
by_LA_Code.lab.summary.w = by_LA_Code.lab %>% summarise(
 wm.lab = mean(EFFECTSIZE1), 
  sd.lab = sd(EFFECTSIZE1),
  se.lab = sd(EFFECTSIZE1/sqrt(n())),
  lower.lab.w = wm.lab - se.lab, 
  upper.lab.w = wm.lab + se.lab)

ggplot(by_LA_Code.lab.summary.w, aes(LA_Code, wm.lab)) + geom_point(size = 3) + geom_errorbar(aes(ymin = lower.lab.w, ymax= upper.lab.w), width = 0.1) + xlab("LA_Code") + ylab("Weighted Mean Effect Size") + ggtitle("Lab") + ylim(-1,1.5) + geom_hline(yintercept=0, linetype="dashed")
```


